Parsing Postfix log files

### 2007-03-18 doc/logparser.tex
Wrote an outline/structure for the paper, and moved bits around to fit.  I now have an idea of how much progress I've made, yay!

\documentclass[a4paper,12pt,draft]{article}

% Better typesetting of URLs.
\usepackage{url}
\usepackage[final]{graphicx}

@@

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to figure out why
something isn't working, why mail is being rejected, or to check whether
their new anti-spam measures are working.  The more adventurous will have
generated some statistics to show how many hits each of their anti-spam
measures have gotten in the last week, and possibly even generated some
graphs to clearly illustrate the point to management or complaining
users\footnote{This was the author's first real foray into processing
Postfix logs}.  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing needs to be per-connection or per-queueid,
rather than per-line.  One of the barriers to this type of processing is
the unstructured nature of Postfix's logs: each logging line was added on
an ad hoc basis as Wietse found a requirement or added new functionality.
This paper attempts to document the difficult process of parsing Postfix
logs, and presents a program which parses logs and places the resulting
data into a database for further processing.

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent (MTA),  It features extensive anti-spam restrictions,
allowing a mail administrator to deploy those restrictions which they judge
suitable for their needs, rather than a fixed set chosen by Postfix's
author.  These restrictions can be applied and combined on a per-host,
per-recipient or per-sender basis, allowing clients/recipients to be
whitelisted with respect to specific restrictions.  Postfix leverages
simple lookup tables to support arbitrarily complicated user-defined
sequences of checks and exceptions, with the ultimate in flexibility being
the facility to consult an external process which can implement whatever
logic is required; if you wish you can restrict some users to sending mail
on the third Tuesday after pay day only, if that is of any use to you.
Administrators can also supply their own rejection messages, to make it
clear to senders why exactly their mail was rejected.  Unfortunately this
flexibility has a cost: complexity in the logs generated.  While it is easy
to use \texttt{grep(1)} to determine the fate of an individual email, and
even to extend that into generating some useful graphs, following the
progress of an email through Postfix can be quite difficult.  The logs tend
to follow a 90\%/10\% pattern: 90\% of the time the logging is simple, and
the other 10\% of the logging requires 90\% of the code.

This paper and the program it describes are part of a larger project to
optimise your Postfix restrictions, generate statistics and graphs, and
provide a platform on which new restrictions can be trialled and evaluated
to see if they are worth using in the fight against spam.  The program
parses Postfix logs and populates a database with the data gleamed from
those logs, providing a consistent and simple view of your logs which
future tools can utilise.

\section{Parsing details}

The complexity and variation in Postfix's logs requires similar flexibility
in the parser; decoupling the parsing rules from the associated actions
allows the rules to be loaded from an external source without modifications
to the parser source code (significantly lowering the barrier to entry for
new users), and simplifies both parser and rules.  The separation also
allows for a XXX REWRITE separation of functionality: rules handle low
level details of identifying a line, whereas the parser handles the higher
level details of following the path a mail takes through postfix,
assembling the required information, etc.

\subsection{Rules}

@@

    \item [Name] A short name for the rule.

    \item [Description] A longer, more detailed description of the rule.

    \item [Program] The program (smtp, smtpd, qmgr, etc) which the rule
        applies to.  This avoids needlessly trying rules which won't
        match, or worse, might match unintentionally.

    \item [Regex] The regex to compare the log line against.  The regex
        will first have several keywords expanded; this simplifies writing
        rules and avoids needless repetition of complex regexs.

    \item [Result\_cols and connection\_cols] This is how we extract the
        fields in the log line matched by the regex.  The format is:
        \newline 
        hostname = 1; helo = 2; sender = 4; \newline
        i.e. semi-colon separated assignment statements, with the column
        name on the left and the match from the regex (\$1, \$2 etc) on the
        right hand side (without \$).

    \item [Action] The action for the parser to take.

    \item [Queueid] The index of the field in regex giving the queueid.

    \item [Rule\_order] This is used to sort the rules, so that rules
        matching more commonly occurring log lines will be tried first, and
        is updated every time the program is run.

    \item [Priority] This is the user-configurable companion to
        rule\_order: rules with a higher priority will be tried first,
        overriding rule\_order, allowing more specific rules to take
        precedence over more general rules.

    \item [Result]  The result the log line represents: rejection,
        acceptance, info, etc.

    \item [Result\_data and connection\_data] Sometimes rules need to fake
        a piece of data which isn't present in the log line: the most
        common would be smtp\_code when mail is accepted.  The format is
        the same as result\_cols and connection\_cols.

@@

\subsection{Algorithm}

XXX FLOW CHART GOES HERE

%\includegraphics{logparser-flow-chart.ps}

@@

    \item Mail enters the system via SMTP or local submission.
    \item If the mail is rejected, log all data, and finish with the mail.
    \item Follow the progress of the mail until it's either delivered or
        bounced, log all data, and finish with the mail.

@@

\subsubsection{Complications}

@@

>>>>

% $Id$
\documentclass[a4paper,12pt,draft]{article}

% Better typesetting of URLs.
\usepackage{url}
% Include images
\usepackage[final]{graphicx}

@@

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug out a
problem, explain why mail is being rejected, or to check whether their new
anti-spam measures are working.  The more adventurous will have generated
some statistics to show how many hits each of their anti-spam measures have
gotten in the last week, and possibly even generated some graphs to clearly
illustrate the point to management or complaining users\footnote{This was
the author's first real foray into processing Postfix logs}.  Very few will
have performed in-depth parsing and analysis of their logs, where the
parsing must correlate the log lines per-connection or per-queueid, rather
than processing lines independently.  One of the barriers to this type of
processing is the unstructured nature of Postfix's logs, where each logging
line was added on an ad hoc basis as Wietse found a requirement or added
new functionality, and further complication comes from user defined log
messages.  This paper attempts to document the difficult process of parsing
Postfix logs, and presents a program which parses logs and places the
resulting data into a database for further processing.

@@

\subsection{Postfix background}

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive anti-spam restrictions,
allowing a mail administrator to deploy those restrictions which they judge
suitable for their needs, rather than a fixed set chosen by Postfix's
author.  These restrictions can be applied and combined on a per-host,
per-recipient or per-sender basis, allowing clients/recipients to be
whitelisted with respect to specific restrictions.  Postfix leverages
simple lookup tables to support arbitrarily complicated user-defined
sequences of restrictions and exceptions, with the ultimate in flexibility
being the facility to consult an external process which can implement
whatever logic is required: if you wish you can restrict some users to
sending mail on the third Tuesday after pay day only, if that is of any use
to you.  Administrators can also supply their own rejection messages, to
make it clear to senders why exactly their mail was rejected.
Unfortunately this flexibility has a cost: complexity in the logs
generated.  While it is easy to use \texttt{grep(1)} to determine the fate
of an individual email, following the progress of an email through Postfix
can be quite difficult.  The logs tend to follow a 90\%/10\% pattern: 90\%
of the time following the log entries is simple, but the other 10\% of the
logging requires 90\% of the code.

\subsection{Paper background}

This paper and the program it describes are part of a larger project to
optimise your Postfix restrictions, generate statistics and graphs, and
provide a platform on which new restrictions can be trialled and evaluated
to see if they are worth using in the fight against spam.  The program
parses Postfix logs and populates a database with the data gleamed from
those logs, providing a consistent and simple view of your logs which
future tools can utilise.  The program should be reasonably easily
extensible, so if logging to a database doesn't meet your requirements, you
can implement your own processing.

\section{Adaptable parsing: user defined rules}

The complexity and variation in Postfix's logs requires similar flexibility
in the parser; decoupling the parsing rules from the associated actions
allows the rules to be loaded from an external source without modifications
to the parser source code (significantly lowering the barrier to entry for
new users), and greatly simplifies both parser and rules.  The separation
also allows for a clear separation of functionality: rules handle low level
details of identifying a line, whereas the parser handles the higher level
details of following the path a mail takes through postfix, assembling the
required information, etc.

@@

    \item [name] A short name for the rule.

    \item [description] A longer, more detailed description of the rule.

    \item [program] The program (smtp, smtpd, qmgr, etc) which the rule
        applies to.  This avoids needlessly trying rules which won't
        match, or worse, might match unintentionally.

    \item [regex] The regex to compare the log line against.  The regex
        will first have several keywords expanded; this simplifies writing
        rules and avoids needless repetition of complex regex components.

    \item [result\_cols and connection\_cols] This is how we extract the
        fields in the log line matched by the regex.  The format is:
        \newline 
        hostname = 1; helo = 2; sender = 4; \newline
        i.e. semi-colon separated assignment statements, with the column
        name on the left and the match from the regex (\$1, \$2 etc) on the
        right hand side (without \$).

    \item [action] The action for the parser to take.

    \item [queueid] The index of the field in regex giving the queueid.

    \item [rule\_order] This is used to sort the rules, so that rules
        matching more commonly occurring log lines will be tried first, and
        is updated every time the program is run.

    \item [priority] This is the user-configurable companion to
        rule\_order: rules with a higher priority will be tried first,
        overriding rule\_order, allowing more specific rules to take
        precedence over more general rules.

    \item [result]  The result the log line represents: rejection,
        acceptance, info, etc.

    \item [result\_data and connection\_data] Sometimes rules need to fake
        a piece of data which isn't present in the log line: the most
        common would be smtp\_code when mail is accepted.  The format is
        the same as result\_cols and connection\_cols.

@@

\section{Parsing algorithm}

The most substantive part of the parser is the parser al . . . . blah

\subsection{Naive approach}

@@

    \item Mail enters the system via SMTP or local submission.

    \item If the mail is rejected, log all data, and finish with the mail.

    \item Follow the progress of the mail until it's either delivered or
        bounced, log all data, and finish with the mail.

@@

\subsection{Complications encountered}

@@

\subsection{Full algorithm}

\includegraphics{logparser-flow-chart.ps}

\section{Limitations and improvements}

\section{Summary}
